\documentclass[tikz,border=10pt,fleqn]{article}
\title{Broken Strings and Absolute Time}
\author{Hans Halvorson}
\date{\today}
\usepackage{pgfplots}
\pgfplotsset{compat=newest}

%% to do: bibliography
\usepackage{doi}
\usepackage[backend=biber,natbib=true,style=authoryear]{biblatex}
\addbibresource{str.bib}

\usepackage{amsfonts}
\usepackage{amsthm}
\theoremstyle{definition}
\usepackage{amsthm}
\newtheorem*{question}{Question}
\newtheorem*{fact}{Fact}

\newcommand{\vecc}[1]{\overrightarrow{#1}}

\begin{document}

%% don't bury the lede: "the string" is ambiguous!

%% Born rigidity

%% what do Alice and Bob "see"? (Perhaps draw some pictures with their
%% lightcones being dragged along.)

%% Rindler horizon
%% isn't there a situation in which Bob falls outside of Alice's
%% horizon? Yes, I think if Bob starts out at distance more than k,
%% where k is the acceleration

\maketitle

TO DO:
\begin{itemize}
\item Is it correct that Bob's coordinatization of Alice's motion will
  be in the Rindler coordinates?
\item What is Alice's velocity relative to Bob?   
\end{itemize}

\section{Introduction}

In his famous article ``How to teach special relativity'', John Bell
\citeyearpar{bell} claims that a certain physical event $E$ occurs in
a certain scenario.\footnote{As several people have pointed out,
  Bell's thought experiment had already been suggested by a few others
  before him, most notably \citet{dewan1959}. In fact, from the
  earliest days of special relativity it was recognized that there are
  some difficulties with describing extended objects in accelerated
  motion.} He then goes on to argue that the occurence of this event
has some rather interesting implications for the interpretation of the
special theory of relativity. Over the past few decades, several
philosophers have invoked Bell's authority in support of the claim
that special relativity is, strictly speaking, false, or at least
incomplete, i.e.\ just a principle theory in wait of a constructive
underpinning \citep[see][]{brown1999,brown2004}. Or perhaps that is
not really what these philosophers are claiming. It is not completely
clear to me what the upshot of Bell's article is supposed to be.

Here is Bell's description of the relevant scenario.
\begin{quote} Three small spaceships, A, B, and C, drift freely in a
  region of space remote from other matter, without rotation and
  without relative motion, with B and C equidistant from A (Fig. 1).
  On reception of a signal from A the motors of B and C are ignited
  and they accelerate gently (Fig. 2).  Let ships B and C be
  identical, and have identical acceleration programmes. Then (as
  reckoned by an observer in A) they will have at every moment the
  same velocity, and so remain displaced one from the other by a fixed
  distance. Suppose that a fragile thread is tied initially between
  projections from B and C (Fig. 3). If it is just long enough to span
  the required distance initially, then as the rockets speed up, it
  will become too short, because of its need to Fitzgerald contract,
  and must finally break. \end{quote} Most commentators on this
thought experiment agree with Bell's claim that the thread will break
--- despite there being strong disagreement about what the phenomenon
entails vis-a-vis the interpretation of special relativity. My primary
claim about the physics is:
\begin{quote} In the two-rocket scenario, there is no physical object
  that could be called ``the thread''. \end{quote} In fact, quite
generally, no extended physical object can have the feature that all
of its parts undergo the same constant acceleration. If part $B$ and
part $C$ are spacelike separated from each other, and if both are
accelerated via the same vector $\vec{a}$, then their internal clocks
can no longer by syncronized, and there is no single thing that could
be called the mereological sum of $B$ and $C$. (These loose claims
correspond to precise mathematical facts that will be illustrated
later.)

\section{Preliminaries}

We assume that Alice (leading) and Bob (trailing) receive the
instruction to accelerate at the same time (in what was originally
their common reference frame). Bob's trajectory (blue) is given by
\[ b(t) \: = \: (\sinh(t),\cosh(t) ) , \]
in the coordinates of the laboratory frame. Here we use the coordinate
convention $(\mathsf{time},\mathsf{space})$. Similarly, Alice's trajectory
(green) is given by
\[ a(t) \: = \: (\sinh (t),\cosh (t)+1) .\] Since $b(t)=(0,1)+a(t)$,
we have $b'(t)=a'(t)$, where \[ a'(t) \: = \: (\sinh (t),\cosh(t)) ,\]
and
\[ a''(t) \: = \: (\cosh (t),\sinh (t)) .\] In particular,
\[ \| a''(t) \|^2 = \sinh ^2 (t)-\cosh ^2(t) = 1 ,\] for all $t$,
i.e.\ the magnitude of Alice and Bob's accelerations are
constant. Similarly, the magnitude of Alice and Bob's proper
velocities are constant. In contrast, Alice's velocity in the
laboratory frame is $v(s)=\sinh (s)/\cosh (s)$ which is zero at $s=0$
while $\lim _{s\to\infty }v(s)=1$.

In the following diagram, the red vector is tangent to Bob's
trajectory at the point $(\sqrt{2},1)$. The dashed blue line is Bob's
orthogonal spacelike hypersurface when he is located at
$(\sqrt{2},1)$, and the dashed green line is Alice's orthogonal
spacelike hypersurface when she is located at $(\sqrt{2}+1,1)$.

\bigskip \noindent \begin{tikzpicture}
    \begin{axis}[
        axis lines=middle,
        xlabel={$x$},
        ylabel={$t$},
        xlabel style={font=\tiny}, % Makes the x-axis label smaller
        ylabel style={font=\tiny}, 
        xmin=-1, xmax=4,
        ymin=-1, ymax=4,
        grid=both,
        grid style={line width=.1pt, draw=gray!10},
        major grid style={line width=.2pt,draw=gray!50},
        axis equal image,
        legend pos=north west,
        tick label style={font=\tiny}, 
    ]
    % Original hyperbola (positive y values only)
    \addplot[domain=1:4, samples=100, smooth, blue] {sqrt(x^2 - 1)};

    % Shifted hyperbola (positive y values only)
    \addplot[domain=2:4, samples=100, smooth, green] {sqrt((x-1)^2 - 1)};

    % Tangent vector at x=2
    \draw[->, red] (axis cs:{sqrt(2)},1)--++(axis direction cs:1,{sqrt(2)});

    % Calculate slope
    \pgfmathsetmacro{\slope}{1/sqrt(2)}

    % Calculate y-intercept
    \pgfmathsetmacro{\yintercept}{1 - \slope*sqrt(2)}

    % Plot the line
    \draw[blue, dashed] (-1,{\slope*-1 + \yintercept}) -- (4,{\slope*4 + \yintercept}) node[right] {};

    \pgfmathsetmacro{\yinterceptNew}{1 - \slope*(1 + sqrt(2))}

    % Plot the new line
    \draw[green, dashed] (-1,{\slope*-1 + \yinterceptNew}) -- (4,{\slope*4 + \yinterceptNew}) node[right] { };
    \end{axis}
  \end{tikzpicture}

  \bigskip \noindent Each each time $t$, Alice and Bob have the same
  tangent vector, and so they foliate Minkowski spacetime into the
  same family of spatial hypersurfaces. However, Alice and Bob are
  located on different timeslices of the relevant foliation. For
  example, Bob reaches coordinate $(\sqrt{2},1)$ in the lab frame at
  the same time that Alice reaches coordinate
  $(1+\sqrt{2},1)$. [Alice's trajectory results from translating Bob's
  trajectory by the vector $\vecc{b(0)a(0)}$.]  At this time, their
  common foliation consists of lines with slope $1/\sqrt{2}$, but Bob
  lies on the line that has $y$-intercept $0$, while Alice lies on the
  line that has $y$-intercept $-1/\sqrt{2}$.

  \begin{fact} The spacetime distance between Alice and Bob does not
    change. To be more precise, if $a(t)$ represents Alice's position
    at time $t$ and $b(t)$ represents Bob's position at time $t$, then
    $\vecc{b(t)a(t)}$ is a spacelike vector and
    $\| \vecc{b(t)a(t)} \|=1$. \end{fact}

This follows directly from the fact that Alice's trajectory is simply
a constant shift of Bob's trajectory.

%% TO DO: a diagram that shows the evolution of Alice and Bob's
%% lightcones

\newcommand{\drawcone}[2]{% takes two arguments: a and b
  \def\a{#1} % x-coordinate of the vertex
  \def\b{#2} % y-coordinate of the vertex
  % Draw and fill the cone
  \fill[gray, opacity=0.3] (\a,\b) -- (\a-10,\b-10) -- (\a+10,\b-10) -- cycle;
}

\bigskip \noindent \begin{tikzpicture}
    \begin{axis}[
        axis lines=middle,
        xlabel={$x$},
        ylabel={$t$},
        xlabel style={font=\tiny}, % Makes the x-axis label smaller
        ylabel style={font=\tiny}, 
        xmin=-1, xmax=4,
        ymin=-1, ymax=4,
        grid=both,
        grid style={line width=.1pt, draw=gray!10},
        major grid style={line width=.2pt,draw=gray!50},
        axis equal image,
        legend pos=north west,
        tick label style={font=\tiny}, 
    ]
    % Original hyperbola (positive y values only)
    \addplot[domain=1:4, samples=100, smooth, blue] {sqrt(x^2 - 1)};

    % Shifted hyperbola (positive y values only)
    \addplot[domain=2:4, samples=100, smooth, green] {sqrt((x-1)^2 -
      1)};

  \def\t{1.5} % You can change this value

  % Calculate cosh(t) and sinh(t)
  \pgfmathsetmacro\cosht{cosh(\t)}
  \pgfmathsetmacro\sinht{sinh(\t)}

  % Calculate the constant k for the line equation y = x + k
  \pgfmathsetmacro\k{\sinht - \cosht}
  \pgfmathsetmacro\h{\sinht + \cosht}

  % Define starting point (choose an appropriate x-coordinate)
  \def\startX{-2} % You can change this value

  % Define ending point (cosh(t), sinh(t))
  \def\endX{\cosht}
  \def\endY{\sinht}

  \drawcone{1+\cosht}{\sinht} % Example usage


    \end{axis}
  \end{tikzpicture}

\section{Rindler coordinates}

Assume again that Bob's trajectory is given by
$b(t)=(\sinh (t),\cosh (t))$ in the laboratory frame. We have seen
that at each time $t\in\mathbb{R}$, Bob's simultaneity surface is a
line that passes through the origin $(0,0)$. Bob is always spacelike
to the origin, and the spacetime distance of the segment between
$(0,0)$ and $(\sinh(t) ,\cosh (t))$ is always of length $1$. In fact,
the line connecting $(0,0)$ to $(\sinh (t),\cosh (t))$ also passes
through the curve $t\mapsto \alpha (\sinh (t),\cosh (t))$, and the
distance along this line to the curve is $\alpha$.

\begin{figure}[ht]
\begin{tikzpicture}[scale=1]

    % Draw axes
    \draw[->] (0,0) -- (8,0) node[right] {$x$};
    \draw[->] (0,-8) -- (0,8) node[above] {$ct$};

    % Draw hyperbolic curves (Rindler observers' worldlines)
          \draw[blue,thick,domain=-2.75:2.75, smooth, variable=\y] plot ({cosh(\y)}, {sinh(\y)});        

          \draw[gray,domain=-2:2, smooth, variable=\y] plot ({2*cosh(\y)}, {2*sinh(\y)});

          \draw[gray,domain=-1.5:1.5, smooth, variable=\y] plot ({3*cosh(\y)}, {3*sinh(\y)});

         \draw[gray,domain=-1.2:1.2, smooth, variable=\y] plot ({4*cosh(\y)}, {4*sinh(\y)}); 

         \draw[green,thick,domain=-2.6:2.6, smooth, variable=\y] plot ({1+cosh(\y)}, {sinh(\y)});    

      \draw[gray] plot[domain=0:8] (\x, {0.5*\x}) node[anchor=south west] {};

            \draw[gray] plot[domain=0:8] (\x, {0.25*\x}) node[anchor=south west] {};

            \draw[gray] plot[domain=0:8] (\x, {0.75*\x}) node[anchor=south west] {};

                \draw[gray] plot[domain=0:8] (\x, {0.9*\x})
                node[anchor=south west] {};
                

\end{tikzpicture}
\caption{The blue hyperbola is the set of all points that are
  spacelike to the origin and at distance $1$. The green hyperbola is
  the shift of the blue hyperbola one unit to the right. The gray
  hyperbolas are lines of constant distance $n$ to the origin. As is
  clear from the picture, the points on the green hyperbola are
  increasingly distant from the origin. The horizontal gray lines
  correspond to amount of elapsed time along each of the hyperbolas of
  constant distance.}
\end{figure}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:



\section{Conclusion}

\begin{fact} The point $(1,0)=b(0)$ lies on Alice's simultaneity
  surface for all $t\in\mathbb{R}^+$. What's more, $\vecc{b(0)a(t)}$
  is spacelike, and $\| \vecc{b(0)a(t)}\| =1$ for all
  $t\in\mathbb{R}^+$. \end{fact}

\begin{proof} The hyperbola
  $\{ (\cosh (t),\sinh(t)):t\in\mathbb{R} \}$ consists of points that
  are Minkowski distance $1$ from $(0,0)$. Hence the hyperbola
  $\{ (\cosh (t)+1,\sinh (t)):t\in\mathbb{R}\}$ consists of points
  that are Minkowski distance $1$ from $(1,0)$.
\end{proof}


\begin{question} What is the orthogonal (temporal) distance between
  Alice and Bob's hypersurfaces at time $t$? Conjecture: it is the
  same as the length of the tangent vector $v(t)$. To see this, just
  do the pythagorean theorem with the triangle where $v(t)$ is the
  hypotenuse. \end{question}

\begin{question} What's does Alice's motion look like to Bob? What
  does Bob's motion look like to Alice? \end{question}

\begin{question} What is the distance, in Bob's frame, between Bob and
  Alice? \end{question}

\begin{question} Show that $\| a'(t)\|$ is bounded in
  length. \end{question}

\printbibliography 


\end{document}




%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
